A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-02-07 sections/Metatheory.tex
Upto Section 4.4

@@

\begin{figure}[t]
\framebox{$\Gm \sto \Om$} $\Gm$ transfers to $\Om$.
\begin{gather*}
\inferrule*[right=$\mathtt{{\sto}}\Om$]
{~}
{\Om \sto \Om}
\quad
\inferrule*[right=$\mathtt{{\sto}\al}$]
{\Om\vdash\tau \\ \Om,[\tau/\al]\Gm \sto \Om}
{\Om,\al,\Gm \sto \Om}
\end{gather*}
\caption{Transfer Rules}
\label{fig:trans}
\end{figure}

The transfer rules are shown in Figure~\ref{fig:trans}.
Rule $\mathtt{{\sto}\al}$ recursively replace the first existential variable to a well-scoped mono-type
($\Om$, as a declarative worklist, contains no existential variables),
so the algorithmic worklist will finally become a declarative one.
In order to maintain well-formedness,
the substitution should affect all the judgments and term variable bindings in the scope of $\al$.

The transfer $\Gm\sto\Om$ is not a function.
On the one hand, there are infinite possibilities to instantiate an existential variable,
thus a $\Gm$ could be transferred to different $\Om$'s.
On the other hand, any valid mono-type in $\Om$ can be represented by an existential variable in $\Gm$,
thus different $\Gm$'s could be transferred to a single $\Om$.

The following lemmas~\ref{lem:insert} and \ref{lem:extract}
generalize Rule $\mathtt{{\sto}\al}$ from substituting the first existential variable
to substituting any existential variable.

@@

\begin{figure}[t]
\framebox{$\Om \rto \Om'$} $\Om$ reduces to $\Om'$.
\begin{gather*}
\begin{aligned}
\Om,a &\rto \Om \qquad \Om \qquad \Om,x:A \rto \Om\\
\Om\Vdash A\le B &\rto \Om \text{ when } \Om\vdash A\le B\\
\Om\Vdash e\Lto A &\rto \Om \text{ when } \Om\vdash e\Lto A\\
\Om\Vdash e\To_a \jg &\rto \Om\Vdash[A/a]\jg \text{ when } \Om\vdash e\To A\\
\Om\Vdash \appInfAlg{A}{e} &\rto \Om\Vdash[C/a]\jg \text{ when } \Om\vdash \appInf{A}{e}{C}\\
\end{aligned}
\end{gather*}
\caption{Declarative Worklist Typing}
\label{fig:decl:worklist}
\end{figure}

Figure~\ref{fig:decl:worklist} relates the worklist-style declarative system
to the original declarative system.
A successful reduction means a valid variable declaration,
a valid subtyping or checking judgment, or a successful guessing for inferences.
Additionally, judgment chains are unfolded to match the original declarative rules.
Although $\Om$ is defined differently from $\Psi$,
our declarative worklist $\Om$ contains a ordered collection of variable definitions,
which naturally fits declarative subtyping and typing.

@@

The proof proceeds by induction on derivation of $\Gm\redto\nil$.
Proper applications of lemmas \ref{lem:insert} and \ref{lem:extract} finish off the theorem.

@@

We prove completeness by induction on the derivation of $\Om\redto\nil$.
Since the declarative worklist is reduced judgment-by-judgment,
the induction always analyse the first judgment until it is satisfied.
As the algorithmic system introduces existential variables,
a declarative rule may correspond to multiple algorithmic rules,
and thus we analyse each of the possible cases.

For subtyping, algorithmic Rules 10 and 11 worth special treatment.
When the induction reaches the $\mathtt{{\le}{\to}}$ case,
the first judgment is of shape $A_1 \to A_2 \le B_1 \to B_2$.
One of the corresponding algorithmic judgment is $\al \le A \to B$.
However, the case analysis does not imply that $\al$ is fresh in $A$ and $B$,
therefore Rule~10 cannot be applied and the proof get stuck.
The following lemma helps us out in those cases:
the success in declarative subtyping indicates the freshness of $\al$ in $A$ and $B$
in its corresponding algorithmic judgment.
In other words, the declarative system do not accept infinite types,
thus, as an example, we cannot find a mono-type $\tau$ such that $\tau\le 1\to \tau$,
which could be transferred by $\al\le 1\to\al$.

>>>>

This section presents the metatheory of the algorithmic system
presented in Section~\ref{}. We show that three main results hold: 
\emph{soundness}, \emph{completeness} and \emph{decidability}.
All the results have been mechanically formalized and proved in the 
Abella theorem prover.

@@

\begin{figure}[t]
\framebox{$\Gm \sto \Om$} $\Gm$ transfers to $\Om$.
\begin{gather*}
\inferrule*[right=$\mathtt{{\sto}}\Om$]
{~}
{\Om \sto \Om}
\quad
\inferrule*[right=$\mathtt{{\sto}\al}$]
{\Om\vdash\tau \\ \Om,[\tau/\al]\Gm \sto \Om}
{\Om,\al,\Gm \sto \Om}
\end{gather*}
\caption{Transfer rules}
\label{fig:trans}
\end{figure}

To aid formalizing the correspondence between the declarative and
algorithmic systems, we use a relation that transfers algorithmic
worklist judgements into corresponding declarative worklist judgements. In essence this
judgment instantiates every existential variable with a well-formed monotype.

The transfer rules are shown in Figure~\ref{fig:trans}.
Rule $\mathtt{{\sto}\al}$ recursively replaces the first existential variable to a well-scoped monotype.
Note that $\Om$, as a declarative worklist, contains no existential
variables. Therefore by recursively applying this rule, the
algorithmic worklist will finally become a declarative one.
In order to maintain well-formedness,
the substitution should affect all the judgments and term variable bindings in the scope of $\al$.

The transfer $\Gm\sto\Om$ is not a function.\bruno{Do you mean here
  ``is not deterministic''? Would that be more appropriate?}
On the one hand, there are infinite possibilities to instantiate an existential variable,
thus a judment $\Gm$ could be transferred to different $\Om$'s.
On the other hand, any valid monotype in $\Om$ can be represented by an existential variable in $\Gm$,
thus different $\Gm$'s could be transferred to a single $\Om$.

Lemmas~\ref{lem:insert} and \ref{lem:extract}
generalize Rule $\mathtt{{\sto}\al}$ from substituting the first existential variable
to substituting any existential variable.

@@

\begin{figure}[t]
\framebox{$\Om \rto \Om'$} $\Om$ reduces to $\Om'$.
\begin{gather*}
\begin{aligned}
\Om,a &\rto \Om \qquad \Om \qquad \Om,x:A \rto \Om\\
\Om\Vdash A\le B &\rto \Om \text{ when } \Om\vdash A\le B\\
\Om\Vdash e\Lto A &\rto \Om \text{ when } \Om\vdash e\Lto A\\
\Om\Vdash e\To_a \jg &\rto \Om\Vdash[A/a]\jg \text{ when } \Om\vdash e\To A\\
\Om\Vdash \appInfAlg{A}{e} &\rto \Om\Vdash[C/a]\jg \text{ when } \Om\vdash \appInf{A}{e}{C}\\
\end{aligned}
\end{gather*}
\caption{Declarative Worklist Typing.\bruno{I think your names for the
  two relations/definitions aren't great. I think what you're defining
here should be called transfer. The previous relation should probably
be called something different. Maybe something like Declarative
worklist instantiation/realization for the previous relation?}}
\label{fig:decl:worklist}
\end{figure}

Figure~\ref{fig:decl:worklist} relates the worklist-style declarative system
to the original declarative system.
A successful reduction means a valid variable declaration,
a valid subtyping or checking judgment, or a successful guessing for inferences.
Additionally, judgment chains are unfolded to match the original declarative rules.
Although $\Om$ is defined differently from $\Psi$,
our declarative worklist $\Om$ contains a ordered collection of variable definitions,
which naturally fits declarative subtyping and typing.
\bruno{I think that to make things precise what you want to show is
  that you can always define a erasure function that drops the
  judgements from the worklist, and retains only the variable
  declarations (thus becomes a declarative environment). Isn't this
  what you do in your formalization?}

@@

The proof proceeds by induction on derivation of $\Gm\redto\nil$.
Proper applications of lemmas \ref{lem:insert} and \ref{lem:extract}
finish off the theorem.
\bruno{Give a bit more informal explanation here. Something like: If
  $\Gm$ is reducible then we can show that there is always an $\Om$
  where all the corresponding declarative judgements hold.}
\bruno{I wonder if 1 or 2 examples (concrete instantiations of the
  theorem) would be helpful to complement the explanation.}

@@

We prove completeness by induction on the derivation of $\Om\redto\nil$.
Since the declarative worklist is reduced judgment-by-judgment,
the induction always analyses the first judgment until it is satisfied.
As the algorithmic system introduces existential variables,
a declarative rule may correspond to multiple algorithmic rules,
and thus we analyse each of the possible cases.

For subtyping, algorithmic Rules 10 and 11 require special treatment.
When the induction reaches the $\mathtt{{\le}{\to}}$ case,
the first judgment is of shape $A_1 \to A_2 \le B_1 \to B_2$.
One of the corresponding algorithmic judgment is $\al \le A \to B$.
However, the case analysis does not imply that $\al$ is fresh in $A$ and $B$,
therefore Rule~10 cannot be applied and the proof gets stuck.
The following lemma helps us out in those cases:
the success in declarative subtyping indicates the freshness of $\al$ in $A$ and $B$
in its corresponding algorithmic judgment.
In other words, the declarative system does not accept infinite types,
thus, as an example, we cannot find a monotype $\tau$ such that $\tau\le 1\to \tau$,
which could be transferred by $\al\le 1\to\al$.
